---
title: "Demultiplexing FastQ Files in the Command Line"
author: Tamas Schauer
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
    html_document:
        toc: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Quick Intro to Command Line

* Open the Terminal app from /Applications/Utilities on your Mac
* Check the following commands

```{bash, comment=NA, message=FALSE, warning=FALSE, results='hide'}
# change to home directory
cd ~

# print working directory
pwd

# make new directory
mkdir Software

# list files
ls -l

# manual page of cp command
man cp
```


## Install JE demultiplexer

* JE is available on github
* download and extract the files

```{bash, comment=NA, message=FALSE, warning=FALSE, results='hide'}
# change to Software directory
cd ~/Software

# download from github
git clone https://github.com/gbcs-embl/Je.git

# copy given version to current directory (.)
cp Je/dist/je_2.0.RC.tar.gz .

# extract
tar xzf je_2.0.RC.tar.gz

# remove the rest (caution: -rf option removes recursively everything without asking!)
rm -rf Je
```

```{bash, comment=NA, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
# check JE manual main page
~/Software/je_2.0.RC/je 

# manual for demultiplex-illu
~/Software/je_2.0.RC/je demultiplex-illu
```


## Setup Project Directory

* create directory structure
* copy fastq files

```{bash, comment=NA, message=FALSE, warning=FALSE, results='hide'}
# start from the home directory
cd ~

# create Projects and Project01 sub-directory
mkdir -p Projects/Project01

# change to Project01 directory
cd Projects/Project01

# create a directory for fastq files
mkdir FastQ

# copy fastq files from Downloads (or where they are) to FastQ directory 
# Note that asterisk is a "wildcard" (any number of unknown characters)
cp ~/Downloads/test_R*.fastq.gz FastQ

# use mv (move files) instead of cp (copy) to avoid redundancy
```

### Barcode File

* create a barcode file in your text editor (my favorite: Atom)
* tab-delimited txt file with 2 columns
* no header in the first line
* place it also into the FastQ directory

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

df <- rbind(
c("sample01", "TTAGGCAT"),
c("sample02", "TGACCAAT"),
c("sample03", "ACAGTGAT"),
c("sample04", "GCCAATAT"))

write.table(df, "~/Projects/Project01/FastQ/test_barcodes.txt", sep="\t", quote = F, col.names = F, row.names = F)

x = kable(df, format="html") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
gsub("<thead>.*</thead>", "", x)
```


```{bash, comment=NA}
# change to FastQ directory
cd ~/Projects/Project01/FastQ 

# list files
ls -l
```

* look into barcode file

```{bash, comment=NA, eval=FALSE}
# cat prints files
cat test_barcodes.txt
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
cat test_barcodes.txt
```



### FastQ Files

* look into R1 file (sample reads)

```{bash, comment=NA, eval=FALSE}
# zcat decompresses and prints files
# | is the pipe symbol (output of zcat goes to the next command)
# head prints the first n lines (-n)
zcat < test_R1.fastq.gz | head -n 12
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
zcat < test_R1.fastq.gz | head -n 12
```


* look into R2 file (index reads)

```{bash, comment=NA, eval=FALSE}
# tail prints the last n lines (-n)
zcat < test_R2.fastq.gz | tail -n 12
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
zcat < test_R2.fastq.gz | tail -n 12
```

* count lines in file

```{bash, comment=NA, eval=FALSE}
# wc (word count) counts lines (-l)
# each read has 4 lines, so the number of reads is 1/4 of lines
zcat < test_R2.fastq.gz | wc -l
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
zcat < test_R2.fastq.gz | wc -l
```


## Run Je

* F1= fastq file for demultiplexing
* BF= barcode text file
* I1= fastq file with barcode reads


```{bash, comment=NA, eval=FALSE}
~/Software/je_2.0.RC/je demultiplex-illu  F1= test_R1.fastq.gz BF= test_barcodes.txt I1= test_R2.fastq.gz
```

```{bash, comment=NA, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
cd ~/Projects/Project01/FastQ 
~/Software/je_2.0.RC/je demultiplex-illu  F1= test_R1.fastq.gz BF= test_barcodes.txt I1= test_R2.fastq.gz
```




```{bash, comment=NA, eval=FALSE}
# list files
ls -l
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
ls -l
```




```{bash, comment=NA, eval=FALSE}
# look at stats output
cat jemultiplexer_out_stats.txt  
```

```{bash, comment=NA, echo=FALSE}
cd ~/Projects/Project01/FastQ 
cat jemultiplexer_out_stats.txt  
```


## Links

* https://gbcs.embl.de/portal/tiki-index.php?page=Je
* https://github.com/gbcs-embl/Je/




